Use of general purpose graphical processing units with MODFLOW-2005

### 2012-08-02 ResponseTo_GW20120518-0099_Review.tex

\documentclass[12pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

\usepackage{geometry} % to change the page dimensions
\geometry{letterpaper} % or letterpaper (US) or a5paper or....
\geometry{margin=1in} % for example, change the margins to 2 inches all round

\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{color}

\usepackage{amsmath}

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim

\begin{document}

\noindent
\textbf{Use of general purpose graphical processing units with MODFLOW-2005 \\ (Ground Water Manuscript GW20120518-0099) }\\

\noindent
\textbf{Authors:} Hughes, J.D., and White, J.T \\

\today \\

\noindent
\textbf{Add some introductory text about how reviewer comments were helpful and have been addressed in the revised manuscript}

Review comments are included (\textit{in italics}) prior to our responses. Revised material in the manuscript are indicated with \textcolor{blue}{blue} text.

\section*{Editor's Comments:}
\begin{enumerate}
\item \textit{Our layout doesn't require section numbering, so please delete those.}

Section numbering has been eliminated in the revised manuscript.

\end{enumerate}

\section*{Reviewer 1 Comments:}
\begin{enumerate}
\item \textit{The introduction should also introduce the speedup potential and bottlenecks of GPGPU and parallel CPU processing for the UPCG solver.}

Response.

\item \textit{Section 2 has too much jargon. I suggest removing what is irrelevant and explaining what is, in terms of how that helps speed up or slow down GPGPU processing compared to serial computing or multi-core processing on CPUs.}

Response.

\item \textit{The manuscript needs to mention how the solver is available for M2K5 for the GPGPU and multi-core CPU versions (are they DLLs, how are they linked into the main code, etc). I do not expect people to generally compile their codes, let alone within CUDA. Some direction therefore is needed on accessibility.}

A section titled ``How to Obtain the Software'' has been added that provides information on how to obtain the software.

\item \textit{Also, direction is needed on hardware that may be best suited for the GPUGPU processing will benefit this manuscript.}

The authors are not experts on the hardware that is best suited for GPGPU processing. Rather than providing guidance on appropriate hardware we feel it is more appropriate for potential users to consult with a qualified software dealer and identify the hardware that is most appropriate for their particular organization.

\item \textit{GPGPU speedup has been compared to the PCG5 solver on a serial CPU. With more advanced solvers coming into the MODFLOW family (for example the XMD solver in MODFLOW-NWT,)it would be interesting to note the speedup of UPCG compared to even more robust higher level ILU solvers which may be even faster than PCG5 for hard problems.}

The analyses presented in this manuscript are not meant to be an exhaustive comparison of GPGPU solutions but an evaluation of the speedup possible.

\item \textit{Issues with GPGPU processing for different schemes are scattered throughout the manuscript. These issues should be collated into a separate section or a table. For instance: GPGPU memory requirements for larger problems; need to maximize the amount of time spent in the linear solvers; the overhead associated with GPGPU-CPU memory copy operations relative to the solution time; higher percentage of runtime spent transferring preconditioner data; why the total time spent applying the preconditioner actually increases for the MILU0 preconditioner; what is less optimized in Intel FORTRAN for multi-core processing; details on problem size limitation due to RAM available on the GPGPU; and others.}

Response.

\end{enumerate}

\section*{Reviewer 2 Comments:}
\begin{enumerate}
\item \textit{Overall, lots of jargon is used. Naturally, this is an emerging field, and one that many readers of Ground Water will not be familiar with, so the use of jargon will be difficult for the uninitiated reader. For example, lines 75-79 are filled with terms that will mean nothing to the average groundwater modeler - consider deleting (or moving to appendix or online material) this paragraph and other similar paragraphs that might not hold much meaning to the average reader.}

Response.

\item \textit{Thorough comparisons were made for different preconditioners and different parallelization strategies and the authors did a great job of making the comparisons fair. Considering the results of the various preconditioners, it appears that Jacobi preconditioning is good enough for many of the problems, which leads me to believe that the K fields are fairly smooth (and the spread of the eigenvalues are nicely clustered). I tend to think of this as the exception in rather than the rule in most practical groundwater models. Can you provide guidance for using MIC vs. POLY vs. Jacobi based on the K field? In other words, I think these results are highly dependent on the K field and only 2 fields are used. How would the results vary if you stretched the contrasts in your K field? In that regard, to be fair, the GMG solver should also be used as a basis for comparison and give insight about which strategy should be used for optimal performance.}

Response.

\item \textit{There seems to be a tradeoff between parallelizing on the CPU vs. the GPU depending on the sophistication of the preconditioner. What is not clear was if it Open-MP was used on PCG2. Reference is made to the work by to Dong and Li, but it appears that this was not considered. How do these results compare to that of Dong and Li? The intent here is not to say that your results are better/worse than those of Dong and Li. Simply, not everyone has access to a video card with double precision GPU capabilities, whereas multicore CPU's seem to be more common, so some guidance of which strategy might be more effective depending on the user's hardware would be helpful (ie, If I were buying a new workstation, should I focus on getting the best multicore CPU, or should I get a less powerful CPU and spend money on a high-end GPU?) I understand that the answer to this question is not straightforward, but any additional insights you could provide regarding this tradeoff would be helpful.}

Response.

\item \textit{Line 26 - "can be *easily* parallelize"}

Response.

\item \textit{Lines 28 - 34 - there is some time discrepancies here from 2005 to 1984, to 1988. Perhaps just say "MODFLOW" on line 28 and remove the "since 1984" and insert "numerous" before "two-and" }

Response.

\item \textit{Line 47 "present" (plural)}

Response.

\item \textit{Line 48 - unclear - faster at what exactly? Operations,cycles/second? }

Response.

\item \textit{Lines 75-108 is filled with a lot of technical jargon. Consider moving some of this to an appendix (as noted above). }

Response.

\item \textit{Line 108 - really, the parallelization you mention can be applied to many simulators, not just MODFLOW-2005. Consider mentioning wider range applicability. }

Response.

\item \textit{Line 213-214 - unclear. Consider rephrasing }

Response.

\item \textit{Line 351 - "developed to *use* the PCG" (one does not "solve" the algorithm) }

Response.

\item \textit{Line376 - as indicated above, can you provide some guidance here about when one approach might be better than another depending on the system properties (size, K field). Also, GMG should be included here to so that the full suite of approaches are compared (weak but highly parallelizable to strong but not very parallelizable). }

Response.

\item \textit{Figure 4 - I think this figure is extremely important as it shows where the speed ups occur. However, it is not clear how much time is spent in the GPU for the matrix-vector and BLAS operations because the bars seem to not show up on the scale? Can you include a value here so the reader does not assume these are zero. }

Response.

\item \textit{In light of the speed ups coming from CUBLAS operations being parallelized over the cells in the model, I am wondering if additional speed up could be obtained by using other primitives in the CUSparse library for the triangular solves and/or the decomposition? Was this considered as well? There is some new work that tries to parallelize ILU (see: http://research.nvidia.com/publication/parallel-incomplete-lu-and-cholesky-factorization-preconditioned-iterative-methods-gpu) }

Response.

\item \textit{As indicated above lines 75-108 are filled with jargon and should be reconsidered. 
Response.}

Response.

\end{enumerate}

\section*{Reviewer 3 Comments:}
\begin{enumerate}
\item \textit{Quality of figures should be improved (all figures but in particular, figures 3 to 6). Texts (titles and legends) are too small to read. Lines are very hard to read.}

Response.

\item \textit{Tables should be reworked. I do not think we need 3 digits after decimal point for these comparisons. Please also consider significant digits in the tables.}

Response.

\item \textit{The authors only consider total execution time which includes time required for both Picard and linear solver iterations. Did you closely examined these iteration performance in solver comparisons? How these iteration performance changes among different preconditioning schemes? }

Response.

\item \textit{Lines 28-32: the authors mentioned computational performance changes due to domain discretization. Does this cause by FD formulation or matrix solver scheme? Explanation is needed. }

Response.

\item \textit{Line 48: 448 $\rightarrow$ 448 cores. }

Response.

\item \textit{Line 114: is L3/L3T correct? }

Response.

\item \textit{Line 123: is L2/T correct? }

Response.

\item \textit{Line 161: What is CUBLAS? }

Response.

\item \textit{Line 199: Problem Description. Reasoning for choosing test cases is helpful for the readers. }

Response.

\item \textit{Line 236: "(10-0)/1000"is not necessary.}

Response.

\end{enumerate}

\end{document}

### 2012-08-27 ResponseTo_GW20120518-0099_InternalReview.tex

\documentclass[12pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

\usepackage{geometry}
\geometry{letterpaper}
\geometry{margin=1in}
\usepackage{graphicx} % support the \includegraphics command and options
\usepackage{color}

\usepackage{amsmath}

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{hyperref}
\renewcommand{\UrlBreaks}{\do\/\do\a\do\b\do\c\do\d\do\e\do\f\do\g\do\h\do\i\do\j\do\k\do\l\do\m\do\n\do\o\do\p\do\q\do\r\do\s\do\t\do\u\do\v\do\w\do\x\do\y\do\z\do\A\do\B\do\C\do\D\do\E\do\F\do\G\do\H\do\I\do\J\do\K\do\L\do\M\do\N\do\O\do\P\do\Q\do\R\do\S\do\T\do\U\do\V\do\W\do\X\do\Y\do\Z}

\begin{document}

\noindent
\textbf{Use of general purpose graphical processing units with MODFLOW-2005 \\ (Ground Water Manuscript GW20120518-0099) }\\

\noindent
\textbf{Authors:} Hughes, J.D., and White, J.T \\

\today \\

\noindent
Internal reviews were performed by (1) Dr. Yong Liu a senior research scientist at the National Center for Supercomputing Applications (NCSA) University of Illinois at Urbana-Champaign, (2) Randy Hanson a research hydrologist in the California Water Science Center, and (3) Vevek Bedekar a professional engineer with  S.S. Papadopulos \& Associates, Inc. Randy Hanson provided a few comments in a review memorandum, which are included and addressed below, and as inline comments in the manuscript pdf; inline comments are addressed in the document titled MODFLOW-GPU\_03.jdhResponse.pdf. Internal review comments were helpful and have been addressed in the revised manuscript. Review comments are included (\textit{in italics}) prior to our responses. Revised material in the manuscript are indicated with \textcolor{blue}{blue} text.

\section*{\small{Yong Liu Comments:}}
\begin{enumerate}
\item \textit{One thing that I would suggest to add is to say a few words about any possible future research (e.g., GPU cluster, Cloud-based GPU resources such as Amazon GPU cluster) and practical usage of GPU solvers for MODFLOW 2005 for real-world use cases (an example real-world use case that would benefit from using this research findings and the GPU solver would be excellent!.}

Response

\item \textit{It would be also nice if the authors can indicate where and how to obtain the source code (will it be in public domain just like the MODFLOW source code?.}

Response

\item \textit{It would be nice if you could also provide the python code so that I or someone else can reproduce some of your experiments.}

Response

\end{enumerate}

\section*{\small{Randy Hanson Comments:}}
\begin{enumerate}
\item \textit{In particular the preconditioning seems to be an issue and one is left wondering if the code could be written better or is it simply a hardware issue given the structure of the code. I have made several inline comments that may give more details as to where these issues might require minor additional elaboration or insight based on your experience. Would a graph showing the change in performance while varying the inner and outer iterations for one of the models provide any additional insight on how to best set the combination of these two, or is this too problem dependent?}

Response.

\item \textit{The reader should also have a clear idea of how he could implement these choices and which choice is best for the problem at hand to some degree (although this may be outside the scope of this paper).}

Response.

\item \textit{The reader may also wonder if additional compilation features such as provided by Intel compiler for memory alignment, vectorization, use of coarrays, use of math-kernal library vectorized functions, or chip specific architectures could provide additional performance enhancements.}

Response.

\item \textit{Inline comments in pdf  MODFLOW-GPU\_03.pdf.}

See responses in MODFLOW-GPU\_03.jdhResponse.pdf.

\end{enumerate}

\section*{Vevek Bedekar Comments:}
\begin{enumerate}
\item \textit{Specify whether this work is first of its kind in applying GPUs in the field of groundwater modeling or specifically in the application of MODFLOW. Otherwise, cite other works that apply GPUs to solve groundwater simulations. The following two search results were found as a result of a quick Google search.}

\textit{\href{http://doi.acm.org/10.1145/1774088.1774588}{Ji, Xiaohui and Cheng, Tangpei and Wang, Qun, 2010, A simulation of large-scale groundwater flow on CUDA-enabled GPUs: Proceedings of the 2010 ACM Symposium on Applied Computing, 2402--2403. doi: 10.1145/1774088.1774588}}

\textit{\href{http://www.gpucomputing.net/?q=node/14182}{Ji, Xiaohui and Cheng, Tangpei and Wang, Qun, 2010, CUDA-based solver for large-scale groundwater flow simulation: Engineering with Computers, 28(1), 13--19. doi: 10.1007/s00366-011-0213-2}}

The second reference (Ji \textit{et al.}, 2012) is a reference in the manuscript but this is difficult to determine from the link provided. The first reference (Ji \textit{et al.}, 2012) was not included in the original manuscript since it is a conference proceeding and may be difficult for some readers to locate. It has been added to ensure Ji \textit{et al.}, 2010 is credited as being the first application of solving MODFLOW using GPGPUs.

\end{enumerate}

\end{document}

### 2012-08-27 ResponseTo_GW20120518-0099_Review.tex

Section numbering has been eliminated in the revised manuscript.

\end{enumerate}

\section*{Reviewer 1 Comments:}
\begin{enumerate}
\item \textit{The introduction should also introduce the speedup potential and bottlenecks of GPGPU and parallel CPU processing for the UPCG solver.}

Additional discussion of ideal parallelization and issues that make achieving ideal speedup has been added to the introduction.

@@

\item \textit{Section 2 has too much jargon. I suggest removing what is irrelevant and explaining what is, in terms of how that helps speed up or slow down GPGPU processing compared to serial computing or multi-core processing on CPUs.}

We have attempted to eliminate all but the essential GPGPU jargon in the main text.  A significant portion of the jargon has been moved to Boxes 1 and 2.

@@

\item \textit{The manuscript needs to mention how the solver is available for M2K5 for the GPGPU and multi-core CPU versions (are they DLLs, how are they linked into the main code, etc). I do not expect people to generally compile their codes, let alone within CUDA. Some direction therefore is needed on accessibility.}

A section titled ``\textbf{How to Obtain the Software}'' has been added that provides information on how to obtain the software.  The source code and a precompiled Windows executable will be provided. Instructions for compiling the source code have been provided in the ``\textbf{How to Obtain the Software}''  section.

@@

\item \textit{Also, direction is needed on hardware that may be best suited for the GPUGPU processing will benefit this manuscript.}

The authors are not experts on the hardware that is best suited for GPGPU processing. Rather than providing guidance on appropriate hardware we feel it is more appropriate for potential users to consult with a qualified software dealer and identify the hardware that is most appropriate for their particular organization.

\item \textit{GPGPU speedup has been compared to the PCG5 solver on a serial CPU. With more advanced solvers coming into the MODFLOW family (for example the XMD solver in MODFLOW-NWT), it would be interesting to note the speedup of UPCG compared to even more robust higher level ILU solvers which may be even faster than PCG5 for hard problems.}

The authors agree with the reviewer that higher level ILU preconditioners may be more effective than the standard MODFLOW PCG solver for particular types of problems. The analyses presented in this manuscript are not meant to be an exhaustive comparison of GPGPU solutions but an evaluation of the speedup possible when compared to a standard MODFLOW solver that uses a similar solution approach.

@@

\item \textit{Issues with GPGPU processing for different schemes are scattered throughout the manuscript. These issues should be collated into a separate section or a table. For instance: GPGPU memory requirements for larger problems; need to maximize the amount of time spent in the linear solvers; the overhead associated with GPGPU-CPU memory copy operations relative to the solution time; higher percentage of runtime spent transferring preconditioner data; why the total time spent applying the preconditioner actually increases for the MILU0 preconditioner; what is less optimized in Intel FORTRAN for multi-core processing; details on problem size limitation due to RAM available on the GPGPU; and others.}

The authors disagree with this comment. We believe the issues related to solution of the groundwater flow equation on a GPGPU are presented in the relevant locations and important issues are summarize again in the ``\textbf{Conclusions}'' section.

@@

\end{enumerate}

\section*{Reviewer 2 Comments:}
\begin{enumerate}
\item \textit{Overall, lots of jargon is used. Naturally, this is an emerging field, and one that many readers of Ground Water will not be familiar with, so the use of jargon will be difficult for the uninitiated reader. For example, lines 75-79 are filled with terms that will mean nothing to the average groundwater modeler - consider deleting (or moving to appendix or online material) this paragraph and other similar paragraphs that might not hold much meaning to the average reader.}

We have tried to eliminate all non-essential GPGPU jargon and make the main text more readable for the typical Ground Water reader.  A significant portion of the jargon has been moved to Boxes 1 and 2 as suggested by the editor.

@@

Response.

\end{enumerate}

\section*{Reviewer 3 Comments:}
\begin{enumerate}
\item \textit{Quality of figures should be improved (all figures but in particular, figures 3 to 6). Texts (titles and legends) are too small to read. Lines are very hard to read.}

The font sizes have been increased on all of the figures. The symbol size and lines on \textbf{figures 3}, \textbf{5}, and \textbf{6} have been increased and converted to solid colored lines, respectively, to improve the readability as suggested.

### 2012-08-27 GW20120518-0099-r01.tex
Latest revisions to document

\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the OpenMP (OpenMP Architecture Review Board, 2005) programming paradigm and observed speedups as large as 5 on machines with multi-core CPUs. Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7. In addition to OpenMP and MPI, GPGPUs also presents an option for linear solver parallelization. In comparison to CPUs, GPGPUs are much faster, have higher bandwidth, and typically more cores (\textit{e.g.}, 448 in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs). Use of GPGPUs to parallelize linear solvers for MODFLOW-2005 is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation. For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems. For solution of the groundwater flow equation, Ji \textit{et al.} (2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.

@@

Although the connectivity of the finite-difference discretization used in MODFLOW-2005 uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row. This storage format is also supported by the CUBLAS libraries. The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (Sadd 2003).

@@

Head-dependent (general head) boundaries were specified on the left and right sides and no flow boundaries at the top and bottom. Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain $\left [ \left(10. - 0.0\right) / 1000. = 0.01 \text{ m/m} \right]$ and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively. For the one layer 1,000 $\times$ 1,000 cell model, conductance and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively. The head-dependent boundaries cause an ambient groundwater flow from left to right. Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d. For multi-layer simulations, the pumping wells withdraw water from the lower model layer. There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries. Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.

@@

The PCG algorithm was implemented for use on GPGPUs using the NVIDIA CUDA-enabled implementation of the BLAS library (CUBLAS). The availability of this API makes it relatively simple to develop parallel GPGPU code. The fork-join parallel model, which uses multiple threads in parallel sections, is used to parallelize code on GPGPUs. Unlike fork-join approaches used on CPUs (OpenMP), which have thread limits determined by the number of CPUs and cores per CPU, GPGPUs have the capability of using thousands to tens of thousands of threads concurrently in parallel sections. The number of concurrent threads that can be used on GPGPUs makes them ideal for parallelization of simultaneous solutions of systems of linear equations like those solved by MODFLOW-2005. The key to successful parallelization on GPGPUs is to (1) use numerical approaches that are highly parallelizable and (2) reduce the need for memory transfers between the CPU and GPGPU as much as possible.

The UPCG solver was developed to solve the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners. An unstructured CSR format was used in the UPCG solver to facilitate use of the CUBLAS library. The UPCG solver also includes options for running sequential and parallel simulations on the CPU. Parallel CPU simulations are executed using the OpenMP programming paradigm.

>>>>

\color{black}Dong and Li (2009) parallelized the MODFLOW PCG solver (Hill 1990) using the OpenMP (OpenMP Architecture Review Board, 2005) programming paradigm and observed speedups as large as 5 on machines with multi-core CPUs. Naff (2008) developed a parallelized linear solver for MODFLOW using the Message Passing Interface (MPI) standard and observed a speedup of 7. In addition to OpenMP and MPI, GPGPUs also presents an option for linear solver parallelization. In comparison to CPUs, GPGPUs are much faster, have higher bandwidth, and typically more cores (\textit{e.g.}, 448 \color{blue} cores \color{black} in the case of the NVIDIA$^{\textregistered}$ Tesla™ C2050 and C2070 GPGPUs). Use of GPGPUs to parallelize linear solvers for MODFLOW-2005 is also attractive because of the availability of the CUDA (Compute Unified Device Architecture) application programming interface (API) for NVIDIA GPGPUs, which supports development of parallel GPGPU code in C/C++ or Python, Fortran, Java, and MATLAB using wrappers around the C/C++ implementation. For example, Soni \textit{et al.} (2012) used the CUDA language to develop a GPGPU code to solve computational fluid dynamics problems. For solution of the groundwater flow equation, Ji \textit{et al.} (2012) developed a parallel GPGPU version of MODFLOW-2000 with a Jacobi preconditioner and observed speedups ranging from 1.8 to 3.7.

@@

Although the connectivity of the finite-difference discretization used in MODFLOW-2005 uses a fixed 5- and 7-point stencil in two- and three-dimensions, respectively, the PCG method in \textbf{algorithm~\ref{pcgalg}} was coded using an unstructured compressed sparse row storage (CSR) format  (Barrett \textit{et al.} 1994) with the diagonal element in the first position in each row. This storage format is also supported by the \color{blue}CUDA-enabled implementation of the BLAS library (CUBLAS). BLAS is a library of Basic Linear Algebra Subprograms\color{black}. The UPCG linear solver includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners (Sadd 2003).

@@

Head-dependent (general head) boundaries were specified on the left and right sides and no flow boundaries at the top and bottom. Head-dependent boundaries are assumed to be located a half model cell away from the edge of the model domain and conductance values were calculated using the average hydraulic conductivity of 4.81 m/d, vertical saturated cell areas, and half the model cell dimensions in the x-direction; head-dependent boundary head values were calculated using the hydraulic gradient over the model domain \color{blue}$\left [ 10. / 1000. = 0.01 \text{ m/m} \right]$ \color{black}and assumed heads at the left and right sides of the model domain of 10.0 and 0.0 m, respectively. For the one layer 1,000 $\times$ 1,000 cell model, conductance and head values of 214.1447 m$^{2}$/d, 160.6085 m$^{2}$/d, 10.025 m, and -0.025 m were specified for the left and right side of the model domain, respectively. The head-dependent boundaries cause an ambient groundwater flow from left to right. Four pumping wells, located in the four cells in the center of the model domain, pump at a constant rate with a total groundwater withdrawal rate of 1,000 m$^3$/d. For multi-layer simulations, the pumping wells withdraw water from the lower model layer. There is no recharge or evapotranspiration; all groundwater flow to the wells is provided by the head-dependent boundaries. Boundary conditions are shown graphically in \textbf{figure~\ref{FigModelDomain}}.

@@

The PCG algorithm was implemented for use on GPGPUs using the NVIDIA \color{blue}CUBLAS library\color{black}. The availability of this API makes it relatively simple to develop parallel GPGPU code. The fork-join parallel model, which uses multiple threads in parallel sections, is used to parallelize code on GPGPUs. Unlike fork-join approaches used on CPUs (OpenMP), which have thread limits determined by the number of CPUs and cores per CPU, GPGPUs have the capability of using thousands to tens of thousands of threads concurrently in parallel sections. The number of concurrent threads that can be used on GPGPUs makes them ideal for parallelization of simultaneous solutions of systems of linear equations like those solved by MODFLOW-2005. The key to successful parallelization on GPGPUs is to (1) use numerical approaches that are highly parallelizable and (2) reduce the need for memory transfers between the CPU and GPGPU as much as possible.

The UPCG solver was developed to \color{blue}use \color{black} the PCG algorithm on GPGPUs and includes Jacobi, ILU0, MILU0, and GLSPOLY preconditioners. An unstructured CSR format was used in the UPCG solver to facilitate use of the CUBLAS library. The UPCG solver also includes options for running sequential and parallel simulations on the CPU. Parallel CPU simulations are executed using the OpenMP programming paradigm.
